The Concurrent Calculi Formalisation Benchmark

### 2024-02-18 short-paper/short-paper.tex
many more notes
some small rewrites to improve readability

In this spirit, we propose a new collection of benchmark problems
specifically designed to address typical issues that arise when
mechanising formal models of concurrent and distributed programming
languages in general, and process calculi in particular.  Concurrent
calculi are notably subtle: for instance it took nine years before the
incorrect proof of type safety in the seminal paper by Honda et
al.~\cite{Honda1998} was fixed in~\cite{Yoshida2007}.  Similarly, many
key results in peer-reviewed papers on session types have turned out
to be false later on~\cite{Gay2020,10.1145/3290343}, illustrating the
value of machine-checkable proofs.

@@

\alb{I really do not like the duplication of the discussion of the
  three issues here and in Sec 2. I understand the desire to highlight
  noted issues, but it does not work. The citations are unconvincing, furthermore}
\paragraph{Linearity.}
In mechanised meta-theory, addressing linearity means choosing an appropriate representation of the linear context.
The crux of the matter is \emph{context splitting}, as showcased by the typing rule for parallel composition.

@@

\paragraph{Scope extrusion.}
The encoding of binding constructs has been extensively studied for type systems based on $\lambda$-calculi, but process calculi present unique challenges.
$\pi$-calculi typically include several different binding constructs: inputs bind a received name or value, recursive processes bind recursion variables, and restrictions bind names.
The first two act similarly to the binders known from $\lambda$-calculi, but restrictions are more challenging due to scope extrusion.

@@

\subsection{Goals}% and how to contribute}
Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
%\alb{We oscillate between concurrency in general and process calculi, here and before}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects involving process calculi?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

@@

Still, these (and other) applications need the basic
techniques that our challenge problems emphasize: as per design goal
\ref{item:goal-reusability}, our problems are drawn from the basic
meta-theory of process calculi (without requiring the development of
new theory), and focus on well-known theorems and results that involve
interesting proof techniques that may be reused in further work.  To
achieve design goal \ref{item:goal-comperison-accessibility}, we have
formulated our challenge problems to explore the three aspects
(linearity, scope extrusion, coinduction) independently, so that they
may be solved individually and in any order; each problem should be
reasonably small and easily understood with basic knowledge of
textbook concurrency theory, process calculi, and type theory.  The
process calculus used in each challenge focuses on the features that
we want to emphasize, and omits all constructs (such as choices) that
would complicate the mechanisation without bringing additional
tangible benefits and insights.  The minimality and uniformity of the
calculi also allows us to target design goal
\ref{item:goal-tutorials}.

@@

\subsubsection{Scope extrusion.}
This is the challenge most closely related to the POPLMark challenge
since it concerns the properties of binders in restrictions and input.

@@

\subsubsection{Coinduction.}
Coinduction is widely supported in proof assistants.
Isabelle, Coq, and Abella seamlessly support coinductive definitions and reasoning.
Agda and Beluga support it through the convenient abstraction of co-patterns~\cite{Abel2013}.
Newer entrants to the field like the Lean proof assistant
are working on adding a notion of coinduction to their system~\cite{Avigad2019,Keizer2023}.

As with most features, proof assistants support
subtly different formalisms that enable different techniques and motivate this challenge.
Some systems provide more than one abstraction to use coinduction.
A good example of this are the parameterized coinduction~\cite{Hur2013}
and interaction tree~\cite{Xia2019} libraries for Coq.

@@

Our coinduction challenge concerns strong barbed bisimulation and
congruence.\alb{we motivate why strong vs weak, but not why barbed}
\lore{Is it not enough to quote the Sangiorgi/Walker book
and go by authority? :) }
Though weak barbed congruence is a more common behavioural equivalence
w.r.t.\ the \(\pi\)-calculus, here we prefer strong equivalences to avoid the
need to abstract over the number of internal transitions, thus
simplifying the theory.  We exclude delegation from the coinduction
challenge since it is orthogonal to our primary aim of exploring
coinductive proof techniques.

The introduction of strong barbed bisimulation
is one of the first steps, when studying the
(observable) behaviour of process calculi, both
in classical textbooks (e.g., \cite{picalcbook})
and in existing 
mechanisations, where we see a variety of
different formalisation choices,
sometimes ad-hoc for the specific
result or development.
As a significant example, we observe that,
while many authors rely
on the native support for coinduction offered by
proof assistants
\cite{Honsell2001,Bengtson2016,Kahsai2008,Thiemann2019,Gay2020},
many others prefer a more ``set-theoretic''
approach
\cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}.

@@

\subsubsection{Evaluation criteria.}
The idea behind our benchmark is to obtain evidence towards answering questions \cref{item:rq1,item:rq2,item:rq3}. We are therefore interested in not only the solutions, but also in the experience of solving the challenges with the chosen approach.
Solutions to our challenges should be compared on three measures:
\begin{enumerate}
\item Mechanisation overhead: the amount of manually written infrastructure and setup needed to express the definitions in the mechanisation;
\item Adequacy of the formal statements in the mechanisation: whether the proven theorems are easily recognisable as the theorems from the challenge; and
\item Cost of entry for the tools and techniques employed: the difficulty of learning to use the techniques.
\end{enumerate}
Solutions to our challenges do not need to strictly follow the definitions and lemmata set out in the problem text, but solutions which deviate from the original challenges must present more elaborate argumentation for their adequacy.

>>>>

In this spirit, we introduce a novel collection of benchmark problems
specifically crafted to tackle common challenges encountered during
the mechanization of formal models for concurrent and distributed
programming languages. Our focus extends particularly to process
calculi.
%
% In this spirit, we propose a new collection of benchmark problems
% specifically designed to address typical issues that arise when
% mechanising formal models of concurrent and distributed programming
% languages in general, and process calculi in particular. 
Concurrent calculi are notably subtle: For instance, it took nine
years before the incorrect proof of type safety in the seminal paper
by Honda et al.~.~\cite{Honda1998} was rectified in Yoshida \&
Vasconcelos’s subsequent work\cite{Yoshida2007}.
% for instance it took nine years before the incorrect proof of type
% safety in the seminal paper by Honda et al was fixed
% in~\cite{Yoshida2007}.
Similarly, many
key results in peer-reviewed papers on session types have turned out
to be false later on~\cite{Gay2020,10.1145/3290343}, illustrating the
value of machine-checkable proofs.

@@

\alb{I really do not like the duplication of the discussion of the
  three issues here and in Sec 2. I understand the desire to highlight
  noted issues, but it does not work. Some citations are unconvincing. Mostly, we are anticipating the description of the challenges. Let's unify this}
\paragraph{Linearity.}
In mechanised meta-theory, addressing linearity means choosing an appropriate representation of the linear context.
The crux of the matter is \emph{context splitting}, as showcased by the typing rule for parallel composition.

@@

\paragraph{Scope extrusion.}
The encoding of binding constructs has been extensively studied
w.r.t.\ the $\lambda$-calculi, but process calculi present additional
challenges.  $\pi$-calculi typically include several different binding
constructs: inputs bind a received name or value, recursive processes
bind recursion variables, and restrictions bind names.  The first two
act similarly to the binders known from $\lambda$-calculi, but
restrictions may be more challenging due to scope extrusion.

@@

\alb{This subsection does not make a lot of sense here -- introductions do not have subsections. Let's move it in ``Design'' }
\subsection{Goals}% and how to contribute}
Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
\alb{Should we stress we are interested in the \textbf{meta-theory} of process calculi, not}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects involving process calculi?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

@@

Still, these (and other) applications need the basic
techniques that our challenge problems emphasize:
Aligned with design goal~\ref{item:goal-reusability}, our problems are rooted in the fundamental meta-theory of process calculi, without necessitating the creation of novel theory. These problems center around well-established theorems and results, showcasing intriguing proof techniques that can be leveraged in future endeavors.
% as per design goal
% \ref{item:goal-reusability}, our problems are drawn from the basic
% meta-theory of process calculi (without requiring the development of
% new theory), and focus on well-known theorems and results that involve
% interesting proof techniques that may be reused in further work. 
To
achieve design goal \ref{item:goal-comperison-accessibility}, we have
formulated our challenge problems to explore the three aspects
(linearity, scope extrusion, coinduction) independently, so that they
may be solved individually and in any order; each problem is %should be
reasonably small and easily understandable with basic knowledge of
textbook concurrency theory, process calculi, and type theory.  The
process calculus used in each challenge focuses on the features that
we want to emphasize, and omits all constructs (such as choices) that
would complicate the mechanisation without bringing additional
tangible benefits and insights.  The minimality and uniformity of the
calculi also allows us to target design goal
\ref{item:goal-tutorials}.

@@

\alb{This needs some work: we do not describe the challenge at all and the citations are not the one I had in mind}
\subsubsection{Scope extrusion.}
This is the challenge most closely related to the POPLMark challenge
since it concerns the properties of binders in restrictions and input.

@@

\subsubsection{Coinduction.}
\alb{I kinda object to ``seamlessly''. In Coq it's a mess, in Abella very very limited, sized types in Agda are shaky. I guess Isabelle is better}

Our coinduction challenge concerns strong barbed bisimulation and
congruence.% \alb{we motivate why strong vs weak, but not why barbed}
% \lore{Is it not enough to quote the Sangiorgi/Walker book
% and go by authority? :)}
Though weak barbed congruence is a more common behavioural equivalence
w.r.t.\ the \(\pi\)-calculus, here we prefer strong equivalences to avoid the
need to abstract over the number of internal transitions, thus
simplifying the theory.  We exclude delegation from the coinduction
challenge since it is orthogonal to our primary aim of exploring
coinductive proof techniques.

The introduction of strong barbed bisimulation is one of the first
steps when studying the (observable) behaviour of process calculi,
both in classical textbooks (e.g., \cite{picalcbook}) and in existing
mechanisations, where we see a variety of different formalisation
choices, sometimes ad-hoc for the specific result or development.

In fact, coinduction is widely supported in proof assistants.
Isabelle, Coq, and Abella seamlessly support coinductive definitions and reasoning.
Agda and Beluga support it through the convenient abstraction of co-patterns~\cite{Abel2013}.
Newer entrants to the field like the Lean proof assistant
are working on adding a notion of coinduction to their system~\cite{Avigad2019}%,Keizer2023}.
%
As with most features, proof assistants support
subtly different formalisms that enable different techniques and motivate this challenge.
Some systems provide more than one abstraction to use coinduction.
A good example of this are the parameterized coinduction~\cite{Hur2013}
and interaction tree~\cite{Xia2019} libraries for Coq.

@@

%As a significant example, we observe that,
While many authors rely on the native support for coinduction offered
by proof assistants
\cite{Honsell2001,Bengtson2016,Kahsai2008,Thiemann2019,Gay2020}, many
others prefer a more ``set-theoretic'' approach
\cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}.
\alb{Actually, Honsell gives up with guarded coinduction and switches to Park's half way}

@@

\alb{This should not be a subsubsection at the same level of coinduction etc}
\subsubsection{Evaluation criteria.}
The idea behind our benchmark is to obtain evidence towards answering
questions \cref{item:rq1,item:rq2,item:rq3}. We are therefore
interested not only in the solutions, but also in the experience of
solving the challenges with the chosen approach.  Solutions to our
challenges should be compared on three axis:
\begin{enumerate}
\item Mechanisation overhead: the amount of manually written infrastructure and setup needed to express the definitions in the mechanisation;
\item Adequacy of the formal statements in the mechanisation: whether the proven theorems are easily recognisable as the theorems from the challenge; and
\item Cost of entry for the tools and techniques employed: the difficulty of learning to use the techniques.
\end{enumerate}
Solutions to our challenges do not need to strictly follow the definitions and lemmata set out in the problem text, but solutions which deviate from the original challenges must present more elaborate argumentation for their adequacy.
